LabMate: a prospectus for types for MATLAB

### 2024-04-11 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] hjux picture
[ writing ] more pics

Under the hood, LabMate translates selected well behaved MATLAB expressions
to typed expressions in its own core theory of types and values.
%
As matrices feature heavily in MATLAB code, the types of matrices
play a central role in this theory.
%
In particular, we acknowledge that the rows and columns of a matrix might pertain to distinct individual entities, rather as the rows ans columns of a spreadsheet do.
%
A matrix type looks like
\[\Matrix{R}{C}{E}{\rs}{\cs}
\]
%
We thus \param{} matrix types by two arbitrary types $R$, for the individuals the rows concern, and $C$ for the columns. We may then give two lists, $\rs = [r_1,\ldots,r_m]\in\List{R}$ (so that each $r_i\in R$) and $\cs = [c_1,\ldots,c_n]\in\List{C}$ of those individuals. A \param{}d type $E(r,c)$ gives the type of entry for any combination of row and column individual, with $E(r_i,c_j)$ being the type of the entry $e_{i,j}$. In the diagram below, the $e$s inside the rectangle are the matrix entries computed by the MATLAB code, while the $r$s and $c$s outside the rectangle are `metadata' coming from the matrix type.
\[\begin{array}{r|ccccc|}
\multicolumn{1}{c}{\;} & c_1      & \cdots & c_j     & \cdots & \multicolumn{1}{c}{c_n}     \\
      \cline{2-6}
r_1    & e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
\vdots &          &        &         &        &         \\
r_i    & e_{i,1}  &        & e_{i,j} &        & e_{i,n} \\
\vdots &          &        &         &        &         \\
r_m    & e_{m,1}  &        & e_{m,j} &        & e_{m,n}  \\
      \cline{2-6}
\end{array}\]
To check that $\Matrix{R}{C}{E}{\rs}{\cs}$ is a valid type, LabMate must:
\begin{enumerate}
\item check that $R$ and $C$ are both valid types
\item check that $E(r,c)$ is a valid type whenever $r\in R$ and $c\in C$
\item check that $\rs\in\List{R}$ and $\cs\in\List{C}$
\end{enumerate}

A notable special case is when both $R$ and $C$ are the unit type $R = C = \One$, that is, the type with exactly one element.
%
$\List{\One}$ amounts to a mere `tally', counting indistinct things, which we allow to be written as a number in decimal notation.
%
In this case, there is only one type of entries $E(r,c)$ (because both $r$
and $c$ must be the unique element of the unit type), and the only
information available in the lists $\rs : \List{\One}$ and
$\cs : \List{\One}$ are their lengths.
%
Thus, we have recovered $\Matrix{\One}{\One}{E}{m}{n}$ as the type of
$m \times n$ matrices with entries of type $E$, where $m$ and $n$ are
the unique lists over the unit type of length $m$ and $n$
respectively.

Matrix types formation can thus be summarised as the inference rule
given in Figure~\ref{fig:matrix_intro}.
%\[
%  \infer{\istype{\Matrix{R}{C}{E}{rs}{\cs}}}
%  {
%    \istype{R}
%    &\quad
%    \istype{C}
%    &\quad
%    x : R, y : C \vdash \istype{E(x,y)}
%    &\quad
%    rs : \List{R}
%    &\quad
%    cs : \List{C}
%  }
%\]
%
\begin{figure*}[th]
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{\istype{\Matrix{R}{C}{E}{\rs}{\cs}}}
      {
        \istype{R}
        &\quad
        \istype{C}
        &\quad
        x : R, y : C \vdash \istype{E(x,y)}
        &\quad
        \rs : \List{R}
        &\quad
        \cs : \List{C}
      }
    \]
  \end{center}
  \caption{Type formation rule for matrix types}
  \label{fig:matrix_intro}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A + B : \Matrix{R}{C}{E}{\rs}{\cs}}
      {
        x : R, y : C \vdash \isadd{E(x, y)}
        &\quad
        A : \Matrix{R}{C}{E}{\rs}{\cs}
        &\quad
        B : \Matrix{R}{C}{E}{\rs}{\cs}
      }
    \]
  \end{center}
  \caption{Addition rule for matrices}
  \label{fig:matrix_add}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A * B : \Matrix{R}{C}{E''}{\rs}{\cs}}
      {
        \ismult{E}{E'}{E''}
        &\quad
        A : \Matrix{R}{C}{E}{\rs}{\ms}
        &\quad
        B : \Matrix{R}{C}{E'}{\ms}{\cs}
      }
    \]
  \end{center}
  \caption{Multiplication rule for matrices}
  \label{fig:matrix_mul}
\end{subfigure}
% FIXME : \caption is broken
\captionwide{Typing rules for matrices}
\label{fig:matrix_rules}
\end{figure*}

>>>>

Under the hood, LabMate translates selected well behaved MATLAB expressions
to typed expressions in its own core theory of types and values.
%
As matrices feature heavily in MATLAB code, the types of matrices
play a central role in this theory.
%
In particular, we acknowledge that the rows and columns of a matrix might pertain to distinct individual entities, rather as the rows ans columns of a spreadsheet do.
%
Internally, a matrix type looks like
\[\Matrix{R}{C}{E}{\rs}{\cs}
\]
%
We thus \param{} matrix types by two arbitrary `header' types $R$, for the individuals the rows concern, and $C$ for the columns. We may then give two lists, $\rs = [r_1,\ldots,r_m]\in\List{R}$ (so that each $r_i\in R$) and $\cs = [c_1,\ldots,c_n]\in\List{C}$ of those individuals. A \param{}d type $E(r,c)$ gives the type of entry for any combination of row and column individual, with $E(r_i,c_j)$ being the type of the entry $e_{i,j}$. In the diagram below, the $e$s inside the rectangle are the matrix entries computed by the MATLAB code. The $r$s and $c$s outside the rectangle are \remph{meta}data coming from the matrix type, known only to LabMate, and not available as data to MATLAB.
\[\begin{array}{r|ccccc|}
\; & c_1      & \cdots & c_j     & \cdots & c_n     \\
\hline
r_1    & e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
\vdots &          &        &         &        &         \\
r_i    & e_{i,1}  &        & e_{i,j} &        & e_{i,n} \\
\vdots &          &        &         &        &         \\
r_m    & e_{m,1}  &        & e_{m,j} &        & e_{m,n}  \\
\hline
\end{array}\]
To check that $\Matrix{R}{C}{E}{\rs}{\cs}$ is a valid type, LabMate must:
\begin{enumerate}
\item check that $R$ and $C$ are both valid types
\item check that $E(r,c)$ is a valid type whenever $r\in R$ and $c\in C$
\item check that $\rs\in\List{R}$ and $\cs\in\List{C}$
\end{enumerate}

A notable special case is when both $R$ and $C$ are the unit type $R = C = \One$, that is, the type with exactly one element.
%
$\List{\One}$ amounts to a mere `tally', counting indistinct things, which we allow to be written as a number in decimal notation.
%
In this case, there is only one type of entries $E(r,c)$ (because both $r$
and $c$ must be the unique element of the unit type), and the only
information available in the lists $\rs : \List{\One}$ and
$\cs : \List{\One}$ are their lengths.
%
Thus, we have recovered $\Matrix{\One}{\One}{E}{m}{n}$ as the type of
$m \times n$ matrices with entries of type $E$, where $m$ and $n$ are
the unique lists over the unit type of length $m$ and $n$
respectively.
%
That is what we render in LabMate directives and responses as
\[
\lsb \;m \;\texttt{x}\; n\; \rsb\; E
\]
as in our \texttt{M :: [ 1 x 5 ] double} in Section~\ref{sec:example}.

So much for what it is to be a valid matrix type: what does it mean to be a valid matrix \remph{in} such a type?
%
Firstly, a singleton entry matrix must have singleton header lists $[r]$ and $[c]$.
\[\begin{array}[b]{r|c|l}
\; & c \\
\cline{1-2}
r  & e\in E(r,c) & \quad \in \Matrix{R}{C}{E}{[r]}{[c]} \\
\cline{1-2}
\end{array} 
\]

\[\begin{array}{l}
\begin{array}{r|c|c|}
  & \cs_A & \cs_b \\
\hline
&&\\
&A \in & B \in \\
\rs &\Matrix{R}{C}{E}{\rs}{\cs_A} & \Matrix{R}{C}{E}{\rs}{\cs_B} \\
&&\\
\hline
\end{array}\\
\;\;\; \in \Matrix{R}{C}{E}{\rs}{(\cs_A\append\cs_B)}
\end{array}\]

%Matrix types formation can thus be summarised as the inference rule given in Figure~\ref{fig:matrix_intro}.
%\[
%  \infer{\istype{\Matrix{R}{C}{E}{rs}{\cs}}}
%  {
%    \istype{R}
%    &\quad
%    \istype{C}
%    &\quad
%    x : R, y : C \vdash \istype{E(x,y)}
%    &\quad
%    rs : \List{R}
%    &\quad
%    cs : \List{C}
%  }
%\]
%
\begin{figure*}[th]
%\begin{subfigure}{\textwidth}
%  \begin{center}
%    \[
%      \infer{\istype{\Matrix{R}{C}{E}{\rs}{\cs}}}
%      {
%        \istype{R}
%        &\quad
%        \istype{C}
%        &\quad
%        x : R, y : C \vdash \istype{E(x,y)}
%        &\quad
%        \rs : \List{R}
%        &\quad
%        \cs : \List{C}
%      }
%    \]
%  \end{center}
%  \caption{Type formation rule for matrix types}
%  \label{fig:matrix_intro}
%\end{subfigure}
%\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A + B : \Matrix{R}{C}{E}{\rs}{\cs}}
      {
        x : R, y : C \vdash \isadd{E(x, y)}
        &\quad
        A : \Matrix{R}{C}{E}{\rs}{\cs}
        &\quad
        B : \Matrix{R}{C}{E}{\rs}{\cs}
      }
    \]
  \end{center}
  \caption{Addition rule for matrices}
  \label{fig:matrix_add}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A * B : \Matrix{R}{C}{E''}{\rs}{\cs}}
      {
        \ismult{E}{E'}{E''}
        &\quad
        A : \Matrix{R}{C}{E}{\rs}{\ms}
        &\quad
        B : \Matrix{R}{C}{E'}{\ms}{\cs}
      }
    \]
  \end{center}
  \caption{Multiplication rule for matrices}
  \label{fig:matrix_mul}
\end{subfigure}
% FIXME : \caption is broken
\captionwide{Typing rules for matrices}
\label{fig:matrix_rules}
\end{figure*}
